|

### 2024-09-21 manuscript.Rmd
Minor edits and changed references to numerical formatting

---
title: |
  | Neural correlates of the uncanny valley effect
  | for robots and hyper-realistic masks 
output:
  bookdown::pdf_document2:
    fig_caption: yes
    toc: no
    keep_tex: yes
  word_document: default
  pdf_document:
    toc: no
  html_document: default
bibliography: references.bib
csl: computers-in-human-behavior.csl
---

[**Author information redacted for blind peer review**]

<!-- author: | -->
<!--   | Shona Fitzpatrick$^1$, Ailish K. Byrne$^2$, Alex Headley$^1$, Jet G. Sanders$^3$, -->
<!--   | Helen Petrie$^4$, Rob Jenkins$^1$ & Daniel H. Baker$^{1,5}$ -->

<!-- 1. Department of Psychology, University of York, York, UK -->
<!-- 2. School of Medicine, Keele University, Newcastle-under-Lyme, Staffordshire, UK -->
<!-- 3. Department of Psychological and Behavioural Science, London School of Economics, London, UK -->
<!-- 4. Department of Computer Science, University of York, York, UK -->
<!-- 5. Corresponding author, email: daniel.baker\@york.ac.uk -->

@@

Neural responses to faces and bodies in general are well-characterised, and there appear to be specialised brain regions devoted to both [reviewed in @Hu2020]. For example, areas of the occipital lobe [@Gauthier2000] and fusiform gyrus [@Kanwisher1997] respond more to faces than non-face stimuli, and sections of extrastriate cortex are responsive to bodies [@Downing2001]. There are also event-related potential (ERP) signals associated with face and body stimuli, though their precise role is still debated [@Thierry2007; @Hong2022]. It seems highly likely that 'uncanny' images will activate these same processes, yet it is unclear whether the sense of unease they produce occurs at bottom-up sensory stages or is modulated by more top-down cognitive factors.

@@

```{r stimexamples, fig.cap="Illustrative stimuli from the same categories as used in Experiments 1 and 2. Row (a) shows a human face, an android and a robot, all against white backgrounds. Row (b) shows a human face, a hyper-realistic silicone mask, and a Halloween mask, all against natural backgrounds. Images were taken from a variety of sources that permit reuse in academic contexts and in most cases were not part of the stimulus set from the experiments. The silicone mask image was taken by the authors, and was used in Experiment 2.", fig.align="center", out.width='80%', echo=FALSE}

@@

A total of 29 participants completed Experiment 1 (12 male, 17 female), and 30 different participants completed Experiment 2 (7 male, 23 female). None of the participants had previously taken part in a study using these stimuli, and all were naïve to the hypotheses and wore their normal optical correction if required. Written informed consent was collected before each experiment began, and all procedures were approved by the Ethics committee of [**redacted for blind peer review**].

@@

In Experiment 1, the stimulus set consisted of a total of 90 images, evenly split between three categories: real faces, human-like robots, and mechanical robots. Images all showed the head and shoulders of the subject, had white backgrounds, and were sourced from the internet. In Experiment 2, the stimulus set (first described by @Sanders2019, but here including additional images) consisted of a total of 296 images, comprising real faces (148 images), people wearing silicone masks (74 images), and people wearing obvious masks of the sort typically worn for carnivals and Halloween celebrations (74 images). The backgrounds of these images were more heterogeneous, and showed the natural surroundings of the subject. In both experiments, images involved examples of both genders, and of varied ethnic backgrounds.

@@

After the EEG experiment, participants also completed a series of questionnaires using the Qualtrics platform. These involved rating their perception of a subset of the stimuli (8 from each category), using items from the Godspeed questionnaire [@Bartneck2009]. Inspection time was unlimited. Participants also provided demographic information (age, gender) and completed the GAToRS [@Koverola2022] and AQ [@Baron-Cohen2001] questionnaires, however the results of these additional questionnaires are not presented here.

@@

We performed univariate analyses by conducting Bayesian t-tests [@Rouder2009] between ERPs from pairs of conditions at each time point using a JZS prior, using signals pooled across electrodes P6, P8, PO6 and PO8. The resulting Bayes factor score is a summary of the evidence in favour of either the null hypothesis (that the waveforms are equal) or the alternative hypothesis (that they differ). We use the heuristics proposed by @Jeffreys1961 that Bayes factors >3 ($log_{10}BF_{10}$ > 0.5) constitute some evidence supporting the alternative hypothesis, factors >10 ($log_{10}BF_{10}$ > 1) constitute strong evidence, and factors >30 ($log_{10}BF_{10}$ > 1.5) constitute very strong evidence.

Multivariate pattern analysis was conducted by training a linear support vector machine algorithm [LibSVM, @Chang2011] to discriminate between patterns of activity across electrodes at a specific time point. The patterns came from the human face condition and one of the other conditions, for a single participant. Four examples of each pattern were calculated by averaging over random subsets of 20\% of the available trials from a given condition, and these were used to train the classifier. The accuracy of the classifier was tested on the remaining trials (that were not used in training) for each condition. This process was repeated 1000 times with different trial permutations to obtain an average accuracy, where chance performance is at 50\% correct. The analysis was carried out at all time points, and for each participant separately. We then averaged classifier accuracy across participants, and calculated one sample Bayesian t-tests comparing to chance performance at each time point as described above.

@@

Raw data, processed data, and analysis scripts are freely available through the project repository at: [**redacted for blind peer review**] 
<!-- https://doi.org/10.17605/OSF.IO/5NZ2H -->

@@

```{r RobotsData, fig.cap="Summary of response data and grand mean ERP for Experiment 1. Panel (a) shows d-prime scores for identifying images of human (blue), android (green) and robot (red) faces. Small points show individual participants, and the larger symbols with error bars indicate the group mean and bootstrapped 95\\% confidence intervals. Panel (b) plots reaction times in the same format (note the logarithmic x-axis). Panel (c) shows the grand mean ERP across all participants and conditions, pooled across electrodes P6, P8, PO6 and PO8 (see inset). The shaded region around the curve illustrates the 95\\% confidence interval, and the grey rectangle at the foot indicates the stimulus duration.", fig.align="center", echo=FALSE}

@@

```{r MasksData, fig.cap="Summary of response data and grand mean ERP for Experiment 2. Panel (a) shows d-prime scores for identifying images of human faces (blue), silicone masks (green) and Halloween masks (red). Small points show individual participants, and the larger symbols with error bars indicate the group mean and bootstrapped 95\\% confidence intervals. Panel (b) plots reaction times in the same format (note the logarithmic x-axis). Panel (c) shows the grand mean ERP across all participants and conditions, pooled across electrodes P6, P8, PO6 and PO8 (see inset). The shaded region around the curve illustrates the 95\\% confidence interval, and the grey rectangle at the foot indicates the stimulus duration.", fig.align="center", echo=FALSE}

@@

[**redacted for blind peer review**]
<!-- SF was funded by a YorRobots Venables internship. JGS was funded by the Economic and Social Research Council (Studentship ES/J500215/1) and Prins Bernhard Cultuurfonds (grant E/30/30.13.0630/HVH/IE). Also supported by BBSRC grant BB/V007580/1 awarded to DHB. -->

>>>>

---
title: |
  | Neural correlates of the uncanny valley effect
  | for robots and hyper-realistic masks 
output:
  bookdown::pdf_document2:
    fig_caption: yes
    toc: no
    keep_tex: yes
  word_document: default
  pdf_document:
    toc: no
  html_document: default
bibliography: references.bib
csl: plos.csl
author: |
  | Shona Fitzpatrick$^1$, Ailish K. Byrne$^2$, Alex Headley$^1$, Jet G. Sanders$^3$,
  | Helen Petrie$^4$, Rob Jenkins$^1$ & Daniel H. Baker$^{1,5}$
---

1. Department of Psychology, University of York, York, UK
2. School of Medicine, Keele University, Newcastle-under-Lyme, Staffordshire, UK
3. Department of Psychological and Behavioural Science, London School of Economics, London, UK
4. Department of Computer Science, University of York, York, UK
5. Corresponding author, email: daniel.baker\@york.ac.uk

@@

Neural responses to faces and bodies in general are well-characterised, and there appear to be specialised brain regions devoted to both [reviewed in @Hu2020]. For example, areas of the occipital lobe [@Gauthier2000] and fusiform gyrus [@Kanwisher1997] respond more to faces than non-face stimuli, and sections of extrastriate cortex are responsive to bodies [@Downing2001]. There are also event-related potential (ERP) signals associated with face and body stimuli, though their precise role is still debated [@Thierry2007; @Hong2022]. It seems highly likely that 'uncanny' images will activate these same processes, yet it is unclear whether the initial cause of the sense of unease they produce occurs at bottom-up sensory stages [@MacDorman2024; @MacDorman2016; @Moore2012] or is modulated by more top-down cognitive factors [@Gray2012; @Yam2021].

@@

```{r stimexamples, fig.cap="Illustrative stimuli from the same categories as used in Experiments 1 and 2. Row (a) shows a human face, an android and a robot, all against white backgrounds. Row (b) shows a human face, a hyper-realistic silicone mask, and a Halloween mask, all against natural backgrounds. Images shown here were taken from a variety of sources that permit reuse in academic contexts and in most cases were not part of the stimulus set from the experiments. The silicone mask image was taken by the authors, and was used in Experiment 2.", fig.align="center", out.width='80%', echo=FALSE}

@@

A total of 29 participants completed Experiment 1 (12 male, 17 female), and 30 different participants completed Experiment 2 (7 male, 23 female). Participants were young adults with no history of neurological disorder. None of the participants had previously taken part in a study using these stimuli, and all were naïve to the hypotheses and wore their normal optical correction if required. Written informed consent was collected before each experiment began, and all procedures were approved by the Ethics committee of the Department of Psychology at the University of York.

@@

In Experiment 1, the stimulus set consisted of a total of 90 images, evenly split between three categories: real faces, human-like robots, and mechanical robots. Images all showed the head and shoulders of the subject, had white backgrounds, and were sourced from the Internet. In Experiment 2, the stimulus set (first described by @Sanders2019, but here including additional images) consisted of a total of 296 images, comprising real faces (148 images), people wearing silicone masks (74 images), and people wearing obvious masks of the sort typically worn for carnivals and Halloween celebrations (74 images). The backgrounds of these images were more heterogeneous, and showed the natural surroundings of the subject. In both experiments, images involved examples of both genders, and of varied ethnic backgrounds.

@@

After the EEG experiment, participants also completed a series of questionnaires using the Qualtrics platform. These involved rating their perception of a subset of the stimuli (8 from each category), using items from the Godspeed questionnaire [@Bartneck2009]. Items were selected that were expected to be most closely aligned to measuring the sensation of uncanniness. Inspection time was unlimited. Participants also provided demographic information (age, gender) and completed the GAToRS [@Koverola2022] and AQ [@Baron-Cohen2001] questionnaires, however the results of these additional questionnaires are not presented here.

@@

We performed univariate analyses by conducting Bayesian t-tests [@Rouder2009] between ERPs from pairs of conditions at each time point using a JZS prior, using signals pooled across electrodes P6, P8, PO6 and PO8, which are typically associated with visual responses to faces. The resulting Bayes factor score is a summary of the evidence in favour of either the null hypothesis (that the waveforms are equal) or the alternative hypothesis (that they differ). We use the heuristics proposed by @Jeffreys1961 that Bayes factors >3 ($log_{10}BF_{10}$ > 0.5) constitute some evidence supporting the alternative hypothesis, factors >10 ($log_{10}BF_{10}$ > 1) constitute strong evidence, and factors >30 ($log_{10}BF_{10}$ > 1.5) constitute very strong evidence.

Multivariate pattern analysis was conducted by training a linear support vector machine algorithm [LibSVM, @Chang2011] to discriminate between patterns of activity across electrodes at a specific time point. Previous work has indicated that EEG data do not typically require more complex nonlinear algorithms [@Grootswagers2017]. The patterns came from the human face condition and one of the other conditions, for a single participant. Four examples of each pattern were calculated by averaging over random subsets of 20\% of the available trials from a given condition, and these were used to train the classifier. The accuracy of the classifier was tested on the remaining trials (that were not used in training) for each condition. This process was repeated 1000 times with different trial permutations to obtain an average accuracy, where chance performance is at 50\% correct. The analysis was carried out at all time points, and for each participant separately. We then averaged classifier accuracy across participants, and calculated one sample Bayesian t-tests comparing to chance performance at each time point as described above.

@@

Raw data, processed data, and analysis scripts are freely available through the project repository at: https://doi.org/10.17605/OSF.IO/5NZ2H

@@

```{r RobotsData, fig.cap="Summary of response data and grand mean ERP for Experiment 1. Panel (a) shows d-prime scores for identifying images of human (blue diamonds), android (green squares) and robot (red circles) faces. Small points show individual participants, and the larger symbols with error bars indicate the group mean and bootstrapped 95\\% confidence intervals. Panel (b) plots reaction times in the same format (note the logarithmic x-axis). Panel (c) shows the grand mean ERP across all participants and conditions, pooled across electrodes P6, P8, PO6 and PO8 (see inset). The shaded region around the curve illustrates the 95\\% confidence interval, and the grey rectangle at the foot indicates the stimulus duration.", fig.align="center", echo=FALSE}

@@

```{r MasksData, fig.cap="Summary of response data and grand mean ERP for Experiment 2. Panel (a) shows d-prime scores for identifying images of human faces (blue diamonds), silicone masks (green squares) and Halloween masks (red circles). Small points show individual participants, and the larger symbols with error bars indicate the group mean and bootstrapped 95\\% confidence intervals. Panel (b) plots reaction times in the same format (note the logarithmic x-axis). Panel (c) shows the grand mean ERP across all participants and conditions, pooled across electrodes P6, P8, PO6 and PO8 (see inset). The shaded region around the curve illustrates the 95\\% confidence interval, and the grey rectangle at the foot indicates the stimulus duration.", fig.align="center", echo=FALSE}

@@

SF was funded by a YorRobots Venables internship. JGS was funded by the Economic and Social Research Council (Studentship ES/J500215/1) and Prins Bernhard Cultuurfonds (grant E/30/30.13.0630/HVH/IE). Also supported by BBSRC grant BB/V007580/1 awarded to DHB.
